\textbf{How open are hybrid journals included in transformative agreements?}

### 2024-01-28 analysis/manuscript.tex
Writing methods

---
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    template: apa6.sty
csl: apa.csl
bibliography: references.bib
---

```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.width = 6,
  fig.asp = 0.618,
  out.width = "99%",
  fig.align = "center",
  dpi = 300,
  dev = c("png", "pdf"),
  fig.path = "fig/"
)
options(scipen = 999, digits = 2)
knitr::knit_hooks$set(
  inline = function(x) {
    if (is.numeric(x)) {
      return(prettyNum(x, big.mark = ","))
    } else{
      return(x)
    }
  }
)
```

```{r dependencies}
library(tidyverse, warn.conflicts = FALSE)
library(kableExtra)
```

```{r data, cache=TRUE}
library(hoaddata)

ta_oa_inst <- readr::read_csv(here::here("data", "ta_oa_inst.csv.gz"))

```
# Introduction

# Methods and data

This study combines data from multiple publicly available data sources as visualised in Figure 1. 
Data gathering involved transformative agreement data retrieved from the cOAlition S Journal Checker Tool, which provided information about journal portfolios and participating institutions. 
After identification of hybrid journals, Crossref served as the primary data source for article-level metadata including Creative Commons (CC) license information to determine open access articles on publisher websites.
OpenAlex provided first author affiliation metadata that were subsequently linked to the transformative agreement data to identify open access articles published through these agreements. 
In total, the compiled data set consists of `r hoaddata::jn_ind |> filter(!cr_year %in% c("2017", "2023")) |> distinct(issn_l, cr_year, jn_all) |> pull(jn_all) |> sum()` articles published in `r hoaddata::jn_ind |> filter(!cr_year %in% c("2017", "2023")) |> distinct(issn_l) |> nrow()` hybrid journals between 2018 and 2022.

```{r data_workflow, echo=FALSE, fig.cap="Data collection workflow to obtain article-level transformative agreement data"}
knitr::include_graphics(here::here("analysis", "data_collection_workflow.pdf"))
```

## Transformative agreement and hybrid journal data

```{r journal_oa_exclusion}
oa_journal_status <- readr::read_csv(here::here("data", "oa_journal_status.csv"))

# OA journal total
n_oa_journals <- oa_journal_status %>%
  filter(openalex == TRUE | doaj == TRUE | bielefeld == TRUE) %>% 
  distinct(issn_l) %>%
  nrow()
# By source
#
# GOLD OA
bielefeld <- oa_journal_status %>%
  filter(bielefeld == TRUE) %>% 
  distinct(issn_l)  %>%
  pull()
# DOAJ
doaj <- oa_journal_status %>%
  filter(doaj == TRUE) %>% 
  distinct(issn_l) %>%
  pull()
# OpenAlex venue
openalex <- oa_journal_status %>%
  filter(openalex == TRUE) %>% 
  distinct(issn_l) %>%
  pull()
my_venn <- list(`GOLD OA` = bielefeld, `DOAJ` = doaj, `OpenAlex` = openalex)

p <-
  ggvenn::ggvenn(
    my_venn,
    fill_color = c("grey55",
          "steelblue2",
          "lightsalmon"),
    digits = 0,
    text_color = "black",
    text_size = 3,
    set_name_size = 4
  )
```

Data gathering started with obtaining journals included in transformative agreements using the cOAlition S’ Journal Checker Tool publicly available Transformative Agreement Data dump. 
The dump consisted of multiple online Google spreadsheets where each data file represented one agreement listed in the ESAC Transformative Agreement Registry. 
Each file comprised information about journals and institutions involved per agreement. 
Because the Journal Checker Tool only collected data about active transformative agreements, four different snapshots were used for this analysis:the archived version from July 2021, July 2022, May 2023 and the most current dump downloaded on 11 December 2022. 
This ensured that transformative agreements, which ended in 2021, were also included in the analysis. 
Agreements ending before 2021 could not be obtained.
Overall, the combined Transformative Agreement Data dumps contained `r  length(unique(hoaddata::jct_hybrid_jns$esac_id))` agreements.

@@

Because transformative agreements can include both full open access and hybrid journals, journal data were complemented with information about a journal's open access status using multiple sources: the Directory of Open Access Journals (DOAJ) downloaded on 4 October 2022, OpenAlex (September 2022 dump) and the the Bielefeld list of GOLD OA journals. 
In total, `r n_oa_journals` full open access journals were excluded.
Figure shows a substantial overlap between the three data sources of
`r round(length(unique(Reduce(intersect, my_venn))) / length(unique(Reduce(union, my_venn)))* 100,0)`%.
However, combining different data sources considerably extended the journal matching to exclude full open access journals from transformative agreements.
The Gold OA journals dataset alone adds `r length(unique(Reduce(setdiff, my_venn)))` journals, while the DOAJ comprised `r length(setdiff(my_venn$DOAJ, union(my_venn[["GOLD OA"]], my_venn$OpenAlex)))` full open access journals not listed in either of the other two sources. 
These full open access journals were mostly launched in 2022.

## Article metadata

## Data Analysis

# Results

```{r overview_data_prep}
ta_oa_growth <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  group_by(cr_year) |>
  summarise(ta_oa = n_distinct(doi))
hybrid_oa_growth <- hoaddata::cc_articles |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  group_by(cr_year) |>
  summarise(oa_all = n_distinct(doi))
ta_df <- inner_join(ta_oa_growth, hybrid_oa_growth) |>
  mutate(ta_oa_prop = ta_oa / oa_all) |>
  mutate(no_ta = oa_all - ta_oa, 
         no_ta_prop = 1 - ta_oa_prop) |>
  mutate(cr_year = as.factor(cr_year))
# year all
year_all <-
  hoaddata::jn_ind |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  filter(issn_l %in% hoaddata::cc_articles$issn_l) |>
  distinct(issn_l, cr_year, jn_all) |> 
  group_by(cr_year) |> 
  summarise(n = sum(jn_all))

## Hybrid Journals
 n_hybrid_jns_all <- hoaddata::jn_ind |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  distinct(issn_l) |>
   nrow()

## Hybrid Journals with at least one OA article
n_hybrid_jns <- hoaddata::cc_articles |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  group_by(issn_l) |>
  summarise(oa_all = n_distinct(doi)) |> 
  nrow()

## Hybrid Journals with at least one OA article
n_ta_jns <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  distinct(issn_l) |>
  nrow()

## Hybrid Journals with at least one TA-OA article
n_tas_with_oa <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  distinct(ta_journal_portfolio) |>
  nrow()

## Open access volume
n_oa <-  hoaddata::cc_articles |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  distinct(doi) |>
  nrow()

## Total
n_total <- hoaddata::jn_ind |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  filter(issn_l %in% hoaddata::cc_articles$issn_l) |>
  distinct(issn_l, cr_year, jn_all) |>
  pull(jn_all) |>
  sum()

## TA total
n_tas_with_oa <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  distinct(ta_journal_portfolio) |>
  nrow()

## TA article total
n_ta_oa <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  distinct(doi) |>
  nrow()
```

## Overview

Between 2018 and 2022, a total of `r n_hybrid_jns` out of `r n_hybrid_jns_all` hybrid journals in transformative agreements published at least one open access article under a Creative Commons license.
During this period, these hybrid journals provided open access to `r n_oa` out of `r n_total` articles, representing a five-year open access proportion of `r n_oa / n_total * 100`%.
Authors who could make use of transformative agreements at the time of publication contributed `r n_ta_oa` open access articles to the total.

```{r area_oa_over_the_years}
oa_ta_plot_df <- ta_df |>
  select(cr_year, ta_oa, no_ta) |>
  pivot_longer(cols = c(ta_oa, no_ta), names_to = "type", values_to = "value") |>
  filter(!cr_year %in%  c("2017", "2023")) |>
  inner_join(year_all)
  # plot
oa_ta_plot <-
  ggplot(oa_ta_plot_df, aes(cr_year, value / n, fill = type, group = type)) +
  geom_area(color = "#000000",
            linewidth = 0.15) +
  scale_fill_manual(
    NULL,
    values = c(no_ta = "#b3b3b3a0", ta_oa = "#56B4E9"),
    labels = c(no_ta = "Other", ta_oa = "Agreement"),
    guide = guide_legend(reverse = TRUE)
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.05)),
    labels = scales::percent_format(),
    limits = c(0, .15),
    breaks = c(0, .05, .1, .15)
  ) +
  theme_minimal() +
  labs(y = "Hybrid OA Uptake", x = NULL) +
  theme(
    legend.position = "top",
    legend.justification = "left",
    axis.text.x = element_blank()
  )
```

```{r ta_share_plot}
ta_share_plot <- ta_df |>
  select(cr_year, ta_oa_prop, no_ta_prop) |>
  pivot_longer(cols = c(ta_oa_prop, no_ta_prop), names_to = "type", values_to = "value") |>
  filter(!cr_year %in%  c("2017", "2023"), type == "ta_oa_prop") |>
  ggplot(aes(gsub("^20", "'", as.character(cr_year)), value, group = 1)) +
  geom_bar(stat = "identity", fill = "#56B4E9", color = "#000000", linewidth = 0.15) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), breaks = c(0, 0.25, 0.5),
                     labels = scales::percent) +
  theme_minimal() +
  labs(y = "Percentage\nof Hybrid OA\nvia Agreement", x = "Publication Year")
```

```{r journal_boxplot}
jn_hybrid_oa_growth <- hoaddata::cc_articles |>
  filter(issn_l != "0027-8424") |>
  group_by(cr_year, issn_l) |>
  summarise(oa_all = n_distinct(doi)) |>
  mutate(cr_year = factor(cr_year))
# year all
jn_year_all <-
  hoaddata::jn_ind |>
  filter(issn_l != "0027-8424") |>
  filter(issn_l %in% hoaddata::cc_articles$issn_l) |>
  distinct(issn_l, cr_year, jn_all) |> 
  group_by(cr_year, issn_l) |> 
  summarise(n = sum(jn_all))

oa_jn_box_df <- left_join(jn_hybrid_oa_growth, jn_year_all) |>
  mutate(prop = oa_all / n) |> 
  filter(!cr_year %in%  c("2017", "2023"))
oa_jn_box <- oa_jn_box_df |>
  ggplot(aes(gsub("^20", "'", as.character(cr_year)), prop)) +
  geom_boxplot(outlier.shape = NA) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)),
                     labels = scales::percent) +
  coord_cartesian(ylim = c(0,0.65)) +
  theme_minimal() +
  labs(y = "OA Uptake per Hybrid Journal", x = "Publication Year")
```

```{r}
ta_oa_growth <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424") |>
  group_by(cr_year, issn_l) |>
  summarise(ta_oa = n_distinct(doi))
hybrid_oa_growth <- hoaddata::cc_articles |>
  filter(issn_l != "0027-8424") |>
  group_by(cr_year, issn_l) |>
  summarise(oa_all = n_distinct(doi))
ta_df <- left_join(hybrid_oa_growth, ta_oa_growth) |>
  mutate(across(everything(), .fns = ~replace_na(.,0))) |>
  mutate(no_ta = oa_all - ta_oa) |>
  mutate(cr_year = as.factor(cr_year))

summary_all <- inner_join(ta_df, jn_year_all, by = c("cr_year", "issn_l")) |>
  mutate(prop_ta_oa = ta_oa / n,
         prop_no_ta = no_ta / n) |>
  ungroup() |>
  distinct(cr_year, issn_l, prop_ta_oa, prop_no_ta) |>
  pivot_longer(cols = c(prop_ta_oa, prop_no_ta))
# plot
box_ta_prop <- summary_all |>
  filter(!cr_year %in%  c("2017", "2023")) |>
  # With at least one OA article
  filter(value > 0) |>
  ggplot(aes(gsub("^20", "'", as.character(cr_year)), value, fill = name)) +
  geom_boxplot(outlier.shape = NA, varwidth = TRUE) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)),
                     labels = scales::percent) +
  coord_cartesian(ylim = c(0,0.65)) +
  scale_fill_manual(NULL, values = c(prop_no_ta = "#b3b3b3a0", prop_ta_oa = "#56B4E9"), labels = c(prop_no_ta = "Other", prop_ta_oa = "Agreement"), guide = FALSE) +
  theme_minimal() +
  labs(y = NULL, x = "Publication Year")
```

```{r}
#| fig-cap: "Relative growth of open access in hybrid journals through transformative agreements between 2018 and 2022 per publication year  according to the total number of articles indexed in Crossref. The blue areas represent open access through transformative agreements, as indicated by the affiliation with institutions that had a corresponding transformative agreement in place at the time of publication according to the first author affiliation data in OpenAlex matched with journal and institutional data from cOAlition S Transformative Agreements Public Data. Grey areas depict open access articles where no link to an agreement could be established. (A) Proportion of open access articles in hybrid journals per year. (B) Percentage of hybrid open access via agreements per year. Boxplots show the proportion of open access articles by individual hybrid journals (C) and individual open access uptake rates by individual hybrid journals and open access funding (D) per publication year. Horizontal lines correspond to the lower quartile, median, and upper quartile, with whiskers extending to ±1.5 of interquartile range. The individual outliers are not shown. Note that data on transformative agreements ending before June 2021 were not available at the time of this study."

library(patchwork)
layout = "
EEE
ACD
BCD"
wrap_plots(
  E = guide_area(),
  A = oa_ta_plot,
  B = ta_share_plot,
  C = oa_jn_box,
  D = box_ta_prop,
  design = layout,
  widths = c(0.4, 0.3, 0.3),
  heights = c(0.2, 1, .5),
  guides = "collect") &
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.direction = "horizontal",
    panel.grid.minor = element_blank(),
                panel.border = element_rect(color = "grey50", fill = NA),
    axis.title=element_text(size = 10)) &
    plot_annotation(tag_levels = 'A')
```

```{r}
# in text data
oa_by_year <- oa_ta_plot_df |>
  group_by(cr_year) |>
  summarise(oa = sum(value),
            articles = n) |>
  mutate(prop = oa / articles * 100) |>
  distinct()
```

Figure 2A shows a moderate growth in the proportion of open access articles in hybrid journals, comparing the overall open access uptake and the impact of transformative agreements on this trend.
Over the five-years period from 2018 to 2022, open access increased from `r oa_by_year |> filter(cr_year == "2018") |> pull(prop)`% (n = `r oa_by_year |> filter(cr_year == "2018") |> pull(oa)`) to `r oa_by_year |> filter(cr_year == "2022") |> pull(prop)`% (n = `r oa_by_year |> filter(cr_year == "2022") |> pull(oa)`). 
At the same time, the total article volume of the investigated journals grew from `r oa_by_year |> filter(cr_year == "2018") |> pull(articles)` in 2018 to `r oa_by_year |> filter(cr_year == "2022") |> pull(articles)` in 2022.

Figure 2B highlights that the majority of open access articles in hybrid journals were made available through transformative agreements in 2021 and 2022, contributing `r oa_ta_plot_df |> filter(cr_year == "2022", type == "ta_oa") |> pull(value) / oa_by_year |> filter(cr_year == "2022") |> pull(oa) * 100`% of the total open access article volume in 2022.
However, there was also a notable growth in open access provision through individual publication fees, which increased from `r oa_ta_plot_df |> filter(cr_year == "2018", type == "no_ta") |> mutate(prop = value / n * 100) |> pull(prop)`% (n = `r oa_ta_plot_df |> filter(cr_year == "2018", type == "no_ta") |> pull(value)`) in 2018 to  `r oa_ta_plot_df |> filter(cr_year == "2022", type == "no_ta") |> mutate(prop = value / n * 100) |> pull(prop)`% (n = `r oa_ta_plot_df |> filter(cr_year == "2022", type == "no_ta") |> pull(value)`). This suggests that publishers were able to gain equally from individual and institutional open access publishing options.

Figure 2C depicts the substantial variations among the hybrid journals included in transformative agreements in terms of open access uptake.
Although the median generally follows the trend shown in Figure 2A, the farther stretch of upper quartiles and whiskers over the years illustrates that an increasing number of journals published an above-average proportion of open access articles. 
In 2022, 25% of hybrid journals (n = `r oa_jn_box_df |> filter(cr_year == "2022") |> mutate(top25 = quantile(prop, .75)) |> filter(prop >= top25) |> nrow()`) had an open access uptake of `r oa_jn_box_df |> filter(cr_year == "2022") |> pull(prop) |> quantile(0.75)*100`%, and `r (oa_jn_box_df |> filter(cr_year == "2022", prop >= 0.5) |> nrow()) / length(unique(oa_jn_box_df$issn_l)) *100`% of journals (n = `r oa_jn_box_df |> filter(cr_year == "2022", prop >= 0.5) |> nrow()`) provided the majority of their articles under a Creative Commons license in the same year. 
These journals were, on average, smaller (M = `r oa_jn_box_df |> filter(cr_year == "2022", prop >= 0.5) |> pull(n) |> mean()`, SD = `r oa_jn_box_df |> filter(cr_year == "2022", prop >= 0.5) |> pull(n) |> sd()`) than those with an open access share below 50% (M = `r oa_jn_box_df |> filter(cr_year == "2022", prop <= 0.5) |> pull(n) |> mean()`, SD = `r oa_jn_box_df |> filter(cr_year == "2022", prop <= 0.5) |> pull(n) |> sd()`). 
Notable exception of large journals with an above-average open access proportion were Physical Review D, a high-energy physics journal covered by the SCOAP3 consortium that provided open access to `r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "2470-0010") |> pull(oa_all)` out of `r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "2470-0010") |> pull(n)` articles in 2022, Astronomy and Physics (`r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "0004-6361") |> pull(oa_all)` out of `r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "0004-6361") |> pull(n)` articles in 2022 were open access), which shifted to a subscribe to open business model for all accepted articles as of April 2022, the Journal of Fluid Mechanics (`r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "0022-1120") |> pull(oa_all)` of `r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "0022-1120") |> pull(n)` articles in 2022 were open access) and Bioinformatics (`r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "1367-4803") |> pull(oa_all)` out of `r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "1367-4803") |> pull(n)` articles in 2022 were open access), which flipped to full open access as of January 2023.

When comparing the impact of open access trough transformative agreements across journals, it shows that for many journals these agreements substantially contributed to the growth of open access over the years (Figure 2D). 
Examples of such journals include those with a scope on specific countries or regions, where also transformative agreements were implemented. 
For instance, in 2022, the Germany-based journals Zeitschrift für Erziehungswissenschaft and Zeitschrift für Politikwissenschaft, as well as the Scandinavian Political Studies adressing the Nordic countries, achieved an overall open access uptake of more than 90% just through transformative agreements. 
Despite the rise in transformative agreements, it is worth noting that other means of publishing open access in hybrid journals remained common. 
In total, `r summary_all |> filter(cr_year == "2022", value > 0, name == "prop_no_ta") |> nrow()` journals published open access articles from authors affiliated with institutions without transformative agreements in place, while `r summary_all |> filter(cr_year == "2022", value > 0, name == "prop_ta_oa") |> nrow()` journals published at least one open access article through a transformative agreement in the same year.

## Publishing market

```{r}
jn_ind_jct <- hoaddata::jct_hybrid_jns |>
  inner_join(hoaddata::jn_ind, by = "issn_l") |>
  filter(issn_l != "0027-8424")

ta_active <- ta_oa_inst |>
  filter(issn_l != "0027-8424") |>
  filter(ta_active == TRUE, !is.na(cc)) |>
  group_by(cr_year, issn_l, esac_publisher) |>
  summarise(ta_oa = n_distinct(doi)) |>
  mutate(cr_year = as.character(cr_year))

jn_ind_jct_oa <- jn_ind_jct |>
  filter(!is.na(cc)) |>
  distinct(cr_year, issn_l, esac_publisher, cc, cc_total, jn_all) |>
  group_by(cr_year, issn_l, esac_publisher, jn_all) |>
  summarise(oa = sum(cc_total, na.rm = TRUE))

>>>>

\documentclass[a4paper,man,floatsintext,longtable,noextraspace,12pt]{apa6}

\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{hyperref}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
% make captions italic

% bibliography

% tightlist
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\title{\textbf{How open are hybrid journals included in transformative agreements?}}
\shorttitle{Hybrid OA}
\author{Najko Jahn}
\affiliation{Göttingen State and University Library, University of Göttingen\\
Platz der Göttinger Sieben 1, 37073 Göttingen, Germany\\
najko.jahn@sub.uni-goettingen.de
}

%\authornote{Correspondence concerning this article should be addressed to Najko Jahn}

\abstract{}

\begin{document}
\maketitle

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\hypertarget{methods-and-data}{%
\section{Methods and data}\label{methods-and-data}}

This study combines data from multiple publicly available data sources
as visualised in Figure 1. Data gathering involved transformative
agreement data retrieved from the cOAlition S Journal Checker Tool,
which provided information about journal portfolios and participating
institutions. After identification of hybrid journals, Crossref served
as the primary data source for article-level metadata including Creative
Commons (CC) license information to determine open access articles on
publisher websites. OpenAlex provided first author affiliation metadata
that were subsequently linked to the transformative agreement data to
identify open access articles published through these agreements. In
total, the compiled data set consists of 8,942,669 articles published in
12,858 hybrid journals between 2018 and 2022.

\begin{figure}

{\centering \includegraphics[width=0.99\linewidth]{data_collection_workflow}

}

\caption{Data collection workflow to obtain article-level transformative agreement data}\label{fig:data_workflow}
\end{figure}

\hypertarget{transformative-agreement-and-hybrid-journal-data}{%
\subsection{Transformative agreement and hybrid journal
data}\label{transformative-agreement-and-hybrid-journal-data}}

Data gathering started with obtaining journals included in
transformative agreements using the cOAlition S' Journal Checker Tool
publicly available Transformative Agreement Data dump. The dump
consisted of multiple online Google spreadsheets where each data file
represented one agreement listed in the ESAC Transformative Agreement
Registry. Each file comprised information about journals and
institutions involved per agreement. Because the Journal Checker Tool
only collected data about active transformative agreements, four
different snapshots were used for this analysis:the archived version
from July 2021, July 2022, May 2023 and the most current dump downloaded
on 11 December 2022. This ensured that transformative agreements, which
ended in 2021, were also included in the analysis. Agreements ending
before 2021 could not be obtained. Overall, the combined Transformative
Agreement Data dumps contained 729 agreements.

@@

Because transformative agreements can include both full open access and
hybrid journals, journal data were complemented with information about a
journal's open access status using multiple sources: the Directory of
Open Access Journals (DOAJ) downloaded on 4 October 2022, OpenAlex
(September 2022 dump) and the the Bielefeld list of GOLD OA journals. In
total, 3,439 full open access journals were excluded. Figure shows a
substantial overlap between the three data sources of 72\%. However,
combining different data sources considerably extended the journal
matching to exclude full open access journals from transformative
agreements. The Gold OA journals dataset alone adds 176 journals, while
the DOAJ comprised 10 full open access journals not listed in either of
the other two sources. These full open access journals were mostly
launched in 2022.

\hypertarget{article-metadata}{%
\subsection{Article metadata}\label{article-metadata}}

\hypertarget{data-analysis}{%
\subsection{Data Analysis}\label{data-analysis}}

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{overview}{%
\subsection{Overview}\label{overview}}

Between 2018 and 2022, a total of 11,189 out of 12,857 hybrid journals
in transformative agreements published at least one open access article
under a Creative Commons license. During this period, these hybrid
journals provided open access to 742,369 out of 8,146,958 articles,
representing a five-year open access proportion of 9.1\%. Authors who
could make use of transformative agreements at the time of publication
contributed 328,957 open access articles to the total.

\begin{figure}

{\centering \includegraphics[width=0.99\linewidth]{fig/unnamed-chunk-2-1}

}

\caption{Relative growth of open access in hybrid journals through transformative agreements between 2018 and 2022 per publication year  according to the total number of articles indexed in Crossref. The blue areas represent open access through transformative agreements, as indicated by the affiliation with institutions that had a corresponding transformative agreement in place at the time of publication according to the first author affiliation data in OpenAlex matched with journal and institutional data from cOAlition S Transformative Agreements Public Data. Grey areas depict open access articles where no link to an agreement could be established. (A) Proportion of open access articles in hybrid journals per year. (B) Percentage of hybrid open access via agreements per year. Boxplots show the proportion of open access articles by individual hybrid journals (C) and individual open access uptake rates by individual hybrid journals and open access funding (D) per publication year. Horizontal lines correspond to the lower quartile, median, and upper quartile, with whiskers extending to ±1.5 of interquartile range. The individual outliers are not shown. Note that data on transformative agreements ending before June 2021 were not available at the time of this study.}\label{fig:unnamed-chunk-2}
\end{figure}

Figure 2A shows a moderate growth in the proportion of open access
articles in hybrid journals, comparing the overall open access uptake
and the impact of transformative agreements on this trend. Over the
five-years period from 2018 to 2022, open access increased from 4.3\% (n
= 65,486) to 15\% (n = 249,511). At the same time, the total article
volume of the investigated journals grew from 1,528,051 in 2018 to
1,676,928 in 2022.

Figure 2B highlights that the majority of open access articles in hybrid
journals were made available through transformative agreements in 2021
and 2022, contributing 58\% of the total open access article volume in
2022. However, there was also a notable growth in open access provision
through individual publication fees, which increased from 4.1\% (n =
62,625) in 2018 to 6.3\% (n = 105,896). This suggests that publishers
were able to gain equally from individual and institutional open access
publishing options.

Figure 2C depicts the substantial variations among the hybrid journals
included in transformative agreements in terms of open access uptake.
Although the median generally follows the trend shown in Figure 2A, the
farther stretch of upper quartiles and whiskers over the years
illustrates that an increasing number of journals published an
above-average proportion of open access articles. In 2022, 25\% of
hybrid journals (n = 2,576) had an open access uptake of 29\%, and 6.6\%
of journals (n = 744) provided the majority of their articles under a
Creative Commons license in the same year. These journals were, on
average, smaller (M = 75, SD = 186) than those with an open access share
below 50\% (M = 164, SD = 347). Notable exception of large journals with
an above-average open access proportion were Physical Review D, a
high-energy physics journal covered by the SCOAP3 consortium that
provided open access to 2,341 out of 4,074 articles in 2022, Astronomy
and Physics (1,396 out of 2,230 articles in 2022 were open access),
which shifted to a subscribe to open business model for all accepted
articles as of April 2022, the Journal of Fluid Mechanics (577 of 1,077
articles in 2022 were open access) and Bioinformatics ( out of articles
in 2022 were open access), which flipped to full open access as of
January 2023.

When comparing the impact of open access trough transformative
agreements across journals, it shows that for many journals these
agreements substantially contributed to the growth of open access over
the years (Figure 2D). Examples of such journals include those with a
scope on specific countries or regions, where also transformative
agreements were implemented. For instance, in 2022, the Germany-based
journals Zeitschrift für Erziehungswissenschaft and Zeitschrift für
Politikwissenschaft, as well as the Scandinavian Political Studies
adressing the Nordic countries, achieved an overall open access uptake
of more than 90\% just through transformative agreements. Despite the
rise in transformative agreements, it is worth noting that other means
of publishing open access in hybrid journals remained common. In total,
9,153 journals published open access articles from authors affiliated
with institutions without transformative agreements in place, while
8,780 journals published at least one open access article through a
transformative agreement in the same year.

\hypertarget{publishing-market}{%
\subsection{Publishing market}\label{publishing-market}}

\begin{table}[H]

\caption{\label{tab:unnamed-chunk-5}Hybrid open access through transformative agreements market shares 2018-2022}
\centering
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{lrllrllrllrll}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Hybrid journals} & \multicolumn{3}{c}{Articles} & \multicolumn{3}{c}{OA articles} & \multicolumn{3}{c}{TA OA articles} \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7} \cmidrule(l{3pt}r{3pt}){8-10} \cmidrule(l{3pt}r{3pt}){11-13}
Publisher & Total & \% & C\% & Total & \% & C\% & Total & \% & C\% & Total & \% & C\%\\
\midrule
Elsevier & 1,936 & 17 & 17 & 2,770,826 & 33.8 & 33.8 & 172,723 & 22.9 & 22.9 & 60,440 & 18.3 & 18.3\\
Springer Nature & 2,274 & 20 & 37 & 1,330,430 & 16.2 & 50 & 175,432 & 23.3 & 46.2 & 100,008 & 30.3 & 48.6\\
Wiley & 1,410 & 12.4 & 49.4 & 1,043,052 & 12.7 & 62.7 & 152,723 & 20.3 & 66.5 & 83,443 & 25.3 & 73.9\\
Other & 5,767 & 50.6 & 100 & 3,061,337 & 37.3 & 100 & 252,523 & 33.5 & 100 & 86,294 & 26.1 & 100\\
\bottomrule
\end{tabular}
\end{table}

Analysing hybrid open access across publishers between 2018 and 2022
reveals a large market concentration. Although 48 publishers offered
transformative agreements, the big three commercial publishers Elsevier,
Springer Nature, and Wiley, accounted for 49\% of total article volume
published (see Table ). Together, they published 500,878 open access
articles, representing 66\% of the open access articles in hybrid
journals. Elsevier, Springer Nature, and Wiley made 243,891 articles
open access in hybrid journals through transformative agreements,
resulting in an even larger market share of 74\%. However, there are
differences among the three large publishers. Although Elsevier
published the largest volume of articles (n = 2,770,826, 34\%), it
published a relatively low number of open access articles, including
those that can be associated with transformative agreements. In
contrast, Springer Nature and Wiley provided open access to a larger
proportion of their articles (13\% of Springer Nature articles and 15\%
of Wiley articles were open acccess), leading to higher open access
market shares (23\% Springer Nature resp. 23\% Wiley). This difference
between Elsevier on the one hand and Springer Nature and Wiley on the
other can be attributed to transformative agreements, as the latter made
the majority of their open access articles available through such deals
(Springer Nature 57\% resp. Wiley 55\%).

\begin{center}\includegraphics[width=0.99\linewidth]{fig/unnamed-chunk-6-1} \end{center}

Figure 2 takes a closer look into the growth of hybrid open access
across publishers by year with a focus on open articles enabled by
transformative agreements. Although all publishers show a general
long-term trend towards transformative agreements, Figure 2A and 2B
indicate that, in particular, Wiley's has experienced a substantial
increase in its open access share from 5.9\% (n = 11,628) in 2018 to
26\% (n = 53,503) in 2022, rperesenting an 4.5-fold increase. In
contrast, Elsevier's hybrid journals deonstrated a more modest increase,
from 3.3\% (n = 16,872) in 2018 to 10\% (n = 60,821) in 2022, which is a
relatively low open access share compared to the general trend. In 2018,
Springer Nature had the largest open access proportion among the three
publishers of 8.4\% (n = 19,701), but experienced a relatively slower
growth, resulting in 18\% (n = 52,616) of articles being open access in
Springer Nature hybrid journals in 2022.

The varying degrees of uptake of open access across the three major
publishers can be attributed to distinct approaches to transformative
agreements. Springer Nature, for example, began in 2015 offering
selected consortia, such as the Max Planck Society, the Swedish Bibsam
consortium, and the Finnish FinELib consortium, open access agreements
for its hybrid journal portfolio under the name Springer
Compact\footnote{\url{https://web.archive.org/web/20180414062853id_/http://www.liber2015.org.uk/wp-content/uploads/2015/03/Springer-Compact.pdf}}.
However, these agreements were not included in the data as they
concluded prior to the start of the transformative agreement data
collection in June 2021. Nonetheless, the results suggest the importance
of central agreements for Springer Nature's hybrid open access business
over the past five years (Figure 2B). In 2022, 66\% (n = 34,725) of open
access articles in f Springer Nature hybrid journals were enabled
through transformative agreements. In the same year, 70\% (n = 37,316)
of Wiley's open access articles could be linked to transformative
agreements in 2022. In contrast, Elsevier published fewer than half of
its open access articles through transformative agreements (n = 32,627;
54\%).

The increasing trend towards transformative agreements can be also
observed at the journal-level (Figure 2). While no substantial
differences between open access enabled through transformative
agreements and other revenue source could observed across Elsevier
journals, the distribution of open access across Springer Nature and
Wiley hybrid journals indicates that the growth is not limited to a few
journals, but extends across the portfolio. In particular, Wiley's upper
quantile, which represents the top 25\% of journals in terms of the
proportion of open access articles from transformative agreements,
increased markedly from 13\% in 2020 to 31\% in 2022. At the same time,
the median proportion grew from 7.5\% to 19\%. It is interesting to note
that a small but increasing number of journals from these two publishers
are providing open access to the majority of articles through
transformative agreements. Wiley recorded 68 and Springer Nature 102
hybrid journals with an open access share above 50\% that could be
solely attributed to transformative agreements. Upon inspection, these
journals were mainly society or local language journals with a small
yearly article volume.

\hypertarget{journal-subjects}{%
\subsection{Journal subjects}\label{journal-subjects}}
